---
title: "Sarasota County Census Tract Demographics (2020)"
---

```{r}
#| label: setup
#| message: false
#| echo: false

library(tidyverse)
library(sf)
library(leaflet)
library(htmltools)

# Load census tract data with geometries
sarasota_race <- read_rds("data/sarasota_race.rds")

# Reshape data to wide format to compute demographic percentages by tract
tract_map <- sarasota_race |>
  mutate(pct = round(100 * (value / summary_value), 1)) |>
  pivot_wider(
    id_cols = c(GEOID, NAME, geometry),
    names_from = variable,
    values_from = pct
  ) |>
  st_as_sf() |>
  st_transform(4326)

# Create hover labels summarizing demographic percentages for each tract
labels <- sprintf(
  "<strong>%s</strong><br/>
   Hispanic: %s%%<br/>
   White: %s%%<br/>
   Black: %s%%<br/>
   Asian: %s%%",
  tract_map$NAME,
  tract_map$hispanic,
  tract_map$white,
  tract_map$black,
  tract_map$asian
) |> lapply(HTML)

# Define color scales for each demographic group
pal_hisp  <- colorNumeric("YlOrRd", domain = tract_map$hispanic)
pal_white <- colorNumeric("Blues", domain = tract_map$white)
pal_black <- colorNumeric("Purples", domain = tract_map$black)
pal_asian <- colorNumeric("Greens", domain = tract_map$asian)
```

## Demographic Distribution Across Sarasota County Census Tracts

This map displays racial and ethnic demographics across Sarasota County census tracts from the 2020 Decennial Census:

- **Hispanic**: Higher concentrations in north Sarasota and central tracts.
- **White (non-Hispanic)**: Makes up the majority across coastal and suburban tracts.
- **Black**: Highest shares in Newtown and northern city neighborhoods.
- **Asian**: Distributed across suburban tracts, including Lakewood Ranch and Palmer Ranch.

Use the radio buttons in the upper right to switch between demographic groups. Hover over any census tract to inspect all four percentages.

```{r}
#| label: interactive-map
#| message: false
#| echo: false

leaflet(tract_map, width = "100%", height = "540px") |>
  addTiles() |>
  addPolygons(
    group = "Hispanic %",
    fillColor = ~pal_hisp(hispanic),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels
  ) |>
  addPolygons(
    group = "White %",
    fillColor = ~pal_white(white),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels
  ) |>
  addPolygons(
    group = "Black %",
    fillColor = ~pal_black(black),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels
  ) |>
  addPolygons(
    group = "Asian %",
    fillColor = ~pal_asian(asian),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels
  ) |>
  addLayersControl(
    baseGroups = c("Hispanic %", "White %", "Black %", "Asian %"),
    options = layersControlOptions(collapsed = FALSE)
  )
```
